---
title: "Get Fit with Machine Learning"
author: "Norbert Widmann"
date: "July 19, 2015"
output: html_document
---

<!--  
Has the student submitted a github repo?#
Does the submission build a machine learning algorithm to predict activity quality from activity monitors?
Do the authors describe what they expect the out of sample error to be and estimate the error appropriately with cross-validation?
-->


Executive Summary
-----------------

MotorTrend is giving you all the news regarding the fuel eficiency of automatic transmissions compared with manual transmissions. To answer this question we are using the newest statistical methods to give our readers the best posible information.

The main conclusion is that manual vs. automatic transmission by itself practically does not affect fuel efficiency at all. The Weight of the vehicle is a much more important factor as a predictor for fuel efficiency. So the question "To shift or not to shift" still remains a matter of personal preference. But if you want to safe on fuel better buy a light vehicle.

Data Set
--------

Our experts invested a lot of work to get the best possible data on all the vehicles you might consider buying. This valuable data was stored in the mtcars dataset. You can find a summary of the dataset in the appendix. The transmission is coded in the attribute am. We make a factor with the descriptive names automatic and manual out of it.

```{r echo=FALSE}
mtcars$am[mtcars$am == 0] <- "automatic"
mtcars$am[mtcars$am == 1] <- "manual"
mtcars$am <- factor(mtcars$am)
```

Explorative Data Analysis
-------------------------

As a basis for analysing the dependency between the transmission type and the fuel consumption in detail we look at the data set. You will find a boxplot in the appendix.

From the exploratory box plot we can deduce that there is a dependency between transmission type and fuel consumption. Vehicles with a manuel transmission on average have a better  fuel efficiency. We can also easily see that the variance is different between the two transmission  types and that there are outliers. This is a hint that there may be other possible variables affecting the fuel consumption, e.g. the displacement or the weight of the vehicle. 

We created plots for these two continuous variables in the appendix. From these plots we can deduce that a larger displacement leads to less fuel efficiency and also greater weight leads to less fuel efficient cars. 

As a last step in the explorative analysis we look at the correlation of these attributes:

```{r echo=FALSE}
cordf <- cbind(mtcars$mpg, mtcars$disp, mtcars$wt, mtcars$am)
colnames(cordf) <- c("mpg", "disp", "wt", "am")
cor(cordf)
```

Given the high correlation between weight and displacement we are gonna look only at weight as an additional parameter for the following regression analysis.

Regression Analysis
-------------------

Given the high inverse correlation between vehicle weight and mpg of -0.87 we are first going to look at a simple linear regression of verhicle weight as predictor and mpg as outcome. The plot can be found in the appendix, we added the transmission type as color for informational purposes. Now we look at the coefficients of this regression:

```{r echo=FALSE}
model = lm(mpg ~ wt, mtcars)
summary(model)$coef
```

Every additional ton of vehicle weight leads to a reduced mpg of about 5.34 and therefore causes significantly worse fuel efficiency. The very low P value shows a high statistical significance of vehicle weight as a predictor of fuel efficiency.

Now we add transmission type as a predictor and look at the coefficients again.

```{r echo=FALSE}
model = lm(mpg ~ am + wt, mtcars)
summary(model)$coef
```

From the coefficients we can see that switching from automatic to manual transaction just lowers the medium mpg corrected by weight by about 0.02. Also the P value of of 0.99 shows low statistical significance as a predictor.

Looking at the residual plot of the regression with wt as the predictor we see a reasonable distribution of the residuals. After the regression analysis we recommend that weight is the main predictor of fuel efficiency.

Inference
---------

Finally we test our two models with an anova test to ensure that adding transmission type as a predictor does not improve the model.

```{r echo=FALSE}
model1 <- lm(mpg ~ wt, data=mtcars)
model2 <- lm(mpg ~ am + wt, data=mtcars)
anova(model1, model2)
```

The very high P-value shows again that we do not get improved prediction quality by adding transmission type. Therefore we conclude that weight is by far the better predictor for fuel efficiency.

\newpage

Appendix
--------

The mtcars dataset.

```{r}
str(mtcars)
```

Box plot of transmission type vs. mpg:

```{r}
boxplot(mpg ~ am, mtcars)
```


Plot of displacement vs. mpg:

```{r}
plot(mpg ~ disp, mtcars)
```

Plot of weight of vehicle vs. mpg:

```{r}
plot(mpg ~ wt, mtcars)
```

Plot of linear regression with weight as predictor and mpg as outcome. Vehicles with manual transmissions are red, vehicles with automatic transmission black:

```{r}
plot(mpg ~ wt, mtcars, col=am)
abline(lm(mpg ~ wt, mtcars))
```

Residual plot of the regression with weight as predictor, transmission as color (see above):

```{r}
plot(mtcars$wt, resid(lm(mpg ~ wt, mtcars)), col=mtcars$am)
abline(0, 0)
```

